Machine Learning Tools for Open Cluster Characterization with Gaia DR2 Data

### 2021-01-09 thesis/thesis.tex
thesis: Review Method

Figure \ref{fig:raw_pm_melotte_22} shows \emph{proper motions in right ascension and declination} for a sample of the downloaded dataset for Melotte 22.
At first sight, two main clusters can be distinguished, one of them centered nearly at (0, 0) and the second cluster with center at (20, -45). This second cluster
is the one we are looking for.

However, although this second cluster is almost isolated, there are stars that do not belong to the OC.
Thus, we need more information to properly characterize the OC.

@@

This download is not limited to the size registered in the catalogue.
Instead, a wider region (1.5 times the size of the cluster) is downloaded for each cluster to include stars outside the cluster.
The idea is that the unsupervised model must be able to clusterize the stars that belong to the OC and to discard
those outsiders when characterizing the open cluster.

@@

Taking these considerations into account, the downloaded dataset covers nearly 114 million stars (42GB of compressed data).
This dataset is significantly smaller than the whole Gaia DR2 dataset, which contains information for approximately 1,600 million stars.
However, it is still too large.

@@

After having applied these filters, the number of cluster to analyse is 169 with nearly 75 million stars.

@@

\verb|downloader.Dockerfile| contains the build instructions for the downloader image. We use \href{https://github.com/features/actions}{GitHub Actions}
for automating the process of building a new version of the image with every new push made to the \verb|main| branch.

@@

The second table is partitioned by region, storing all stars related to one region inside an independent table for optimized access performance.
The partition tables are located inside \verb|gaiadr2| schema. See Figure \ref{fig:customdb_diagram}.

@@

PostgreSQL knows that all stars we want are located inside \verb|gaiadr2.region_302| so it does not need to query any other table.
This allows us to download data for as many regions as we want without loosing efficiency when accessing for that data.
% TODO: Revisar corrección de Imanol

@@

The developed code for this work can be found inside the \verb|src| directory at GitHub:
\href{https://github.com/cdalvaro/machine-learning-master-thesis}{cdalvaro/machine-learning-master-thesis} repository.
(For the printed version of this work, you can scan the QR code on cover for accessing to it.)

@@

Parallax is another important feature. It lets us know how far stars are from us.
And, since all stars within an open cluster were born from the same dust cloud, they must all have similar parallax.

@@

First, we correct proper motion in right ascension and declination by dividing them by the parallax.
That way we normalize these quantities and help our clustering models to improve their performance.

Another computed property that we are considering is the modulus of the proper motion.
That way, we relate both features and therefore force our model to keep them tight.

@@

Once we have found the set of features that best describes our problem, we can begin searching for the OC in the study region.

Our first approach to find the open cluster within the region of interest is to use the K-Means algorithm.

@@

As shown in Figure \ref{fig:kmeans_comparisons_melotte_22}, bigger values for the number of clusters allow us to isolate more accurately
the resonance in parallax at $\approx 7.3mas$. However, we have the disadvantage that more groups are formed, while our aim is to find just
one small cluster that contains the stars that belong to the OC and another one with the remaining stars.

This effect complicates our task of finding the desired OC. Therefore, we have to find
a way to set the right value for the number of clusters to isolate the searched cluster without creating too many groups.

@@

K-Means does a good job making an initial clustering, as shown in Figure \ref{fig:kmeans_melotte_22}.
However, too many clusters arise from this characterization and the open cluster is still polluted with stars that do not belong to it.
Moreover, we would like to reduce the amount of clusters too.

@@

Our initial approach allows us to find some groups that potentially contain the desired OC.
However, as we have seen before, the result is not enough accurate.
Therefore, we need a method for merging clusters by migrating stars from one group to another,
so we can end up with the minimum number of clusters while preserving the open cluster.

@@

These encoders are pretrained before fitting the model to generate predictions.
Then, the encoder is used with he aim of transforming input data to the latent space $Z$.
Once the data has been transformed, a K-Means clusterer is used in order to make an initial clustering.

@@

Figure \ref{fig:dec_model_setup} shows the layer setup of our DEC model.
It is simpler than the one tested on the original paper \cite{xie2016unsupervised},
since the number of selected features in our work is smaller.
Therefore, using the same configuration would result in a model so
powerful that would incur in overfitting issues unable to make right predictions.

>>>>

Figure \ref{fig:raw_pm_melotte_22} shows \emph{proper motions in right ascension and declination}
for a sample of the downloaded dataset for Melotte 22.
At first sight, two main clusters can be distinguished, one of them centered nearly at (0, 0)
and the second one with center at (20, -45). This second cluster is the one we are looking for.

However, although the second cluster is almost isolated, there are stars that do not belong to the OC.
Thus, we need more information to properly characterize the open cluster.

@@

This download is not limited to the size registered in the catalogue.
Instead, a wider region (1.5 times the size of the cluster) is downloaded for each cluster to include stars outside the cluster.
The idea is that the unsupervised model must be able to clusterize those stars that belong to the OC
and to discard outsiders when characterizing the open cluster.

@@

Taking these considerations into account, the downloaded dataset covers nearly 114 million stars (around 42GB of compressed data).
This dataset is significantly smaller than the whole Gaia DR2 dataset,
which contains information for approximately 1,600 million stars.
However, it is still too large.

@@

After having applied these filters, the number of cluster to analyze is 169 with nearly 75 million stars.

@@

\verb|downloader.Dockerfile| contains the build instructions for the downloader image.
We use \href{https://github.com/features/actions}{GitHub Actions} for automating the process
for building a new image version with every new push made to the \verb|main| branch.

@@

The second table is partitioned by region,
storing all stars related to one region inside an independent table for optimized access performance.
Partition tables are located inside \verb|gaiadr2| schema. See Figure \ref{fig:customdb_diagram}.

@@

PostgreSQL knows that all stars we want are located inside \verb|gaiadr2.region_302| so it does not need to query any other table.
This allows us to download data for as many regions as we want without loosing performance when accessing for that data.

@@

The developed code for this work can be found inside the \verb|src| directory at GitHub:
\href{https://github.com/cdalvaro/machine-learning-master-thesis}{cdalvaro/machine-learning-master-thesis} repository.
(For the printed version of this work, you can scan the QR code on cover for accessing it.)

@@

Parallax is another important feature. It lets us know how far stars are from us.
In addition, since all stars within an open cluster were born from the same dust cloud,
they must all have similar parallax.

@@

First, we correct proper motion in right ascension and declination by dividing them by the parallax.
That way, we normalize these quantities and help our clustering models to improve their performance.

The modulus of the proper motion is another computed property that we are considering.
We use it to relate both features and therefore, to force our model to keep them tight.

@@

Once we have found the set of features that best describes our problem,
we can begin searching for the OC within the study region.

Our first approach to find the open cluster is using the K-Means algorithm.

@@

As shown in Figure \ref{fig:kmeans_comparisons_melotte_22},
larger values for the number of clusters allow us to isolate more accurately the resonance in parallax at $\approx 7.3mas$.
However, we have the disadvantage that more groups are formed.

This effect complicates our task of finding the desired open cluster,
since we would like to get just two groups, one for the OC and another with the remaining stars.
Therefore, we have to find a way to set the right value for the number of clusters
to isolate the searched cluster without creating too many groups.

@@

K-Means does a good job making an initial clustering,
as shown in Figure \ref{fig:kmeans_melotte_22}.
However, too many clusters arise from this characterization
and the OC is still polluted with stars that do not belong to it.
Moreover, we would like to reduce the amount of clusters too.

@@

Our initial approach allows us to find some groups that potentially contain the desired OC.
However, as we have seen before, the result is not accurate enough.
Therefore, we need a method for merging clusters by migrating stars from one group to another,
so we can end up with the minimum number of clusters while preserving the open cluster.

@@

These encoders are pretrained before fitting the model to generate predictions.
Then, the encoder is used with the aim of transforming input data to the latent space $Z$.
Once the data has been transformed, a K-Means clusterer is used in order to make an initial clustering.

@@

Figure \ref{fig:dec_model_setup} shows the layer setup of our DEC model.
It is simpler than the one tested on the original paper \cite{xie2016unsupervised},
since the number of selected features in our work is smaller than in the original one.
Therefore, using the same configuration would result in a model so
powerful that would incur in overfitting issues unable to make right predictions.
